\pdftitle

### 2017-05-17 maxNt.tex
discussion

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-17T23:04:18+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[final,nonatbib]{nips_2017}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\pdftitle}{Maximum-entropy distributions\\ and representative
  random samples: a dilemma}
\newcommand*{\headtitle}{\pdftitle}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, V. Rostami, E. Torre}

@@

\author{
  V. Rostami\\Hitlerland\\\texttt{v.rostami@fz-juelich.de}
  \And
  P.G.L. Porta Mana\\Mussoliniland\\\texttt{pgl@portamana.org}
  \And
E. Torre\\Moneylaunderingland  
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. The standard \me\ method
\cite{jaynes1954_r1974,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities \cite{***}. But this is only one of its uses. It is
also used for various information-theoretic purposes or to generate
reference probability distributions \cite{***}. In all these uses the \me\
distribution is chosen as the \enquote{maximally noncommittal} one
\cite{jaynes1963}. This adjective means little without further technical
characterizations. Different works give different characterizations, but in
this paper we will use the quoted expression as an umbrella term for all of
them. Our results will not depend on the specific characterization of
\enquote{noncommittal}.

@@

Our discussion pertains to a neurons modelled as binary units at one
specific instant or short window in time. It is possible to similarly
discuss multi-state neuron models and population dynamics; but for
simplicity neurons are here assumed to be in a fixed, active \enquote{$1$}
or inactive \enquote{$0$} state; and evolution, change, time correlations,
and similar concepts do not concern us.

Our notation follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. We intend \enquote{probability} as the degree of
belief that \enquote{would be agreed by all rational men if there were any
  rational men} \cite{good1966}.

@@

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$. A particular sample of $n$
neurons from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$
having fixed binary values $(r_1, \dotsc, r_n)\defs\yr$. We will consider
various averages of the population and the sample. For this purpose we
introduce a general averaging operator $\sav{\dotv}$ defined by
\begin{align}
  \av{\yX}  &\defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
%  &
%  \sav{\yx}  &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
    \\[2\jot]
  \av{\yX \yX} &\defd
\tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3  + \dotsb +  X_{N-1} X_N),
%  &
%    \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3  + \dotsb +  x_{n-1} x_n),
\\[2\jot]
\av{\yX\yX\yX} &\defd
                  \tbinom{N}{3}^{-1}
                  (X_1 X_2 X_3 +% X_1 X_2 X_4
                   \dotsb + X_{N-2} X_{N-1}X_N),
%  &
%    \sav{\yx\yx\yx} &\defd
%                      \tbinom{n}{3}^{-1} (x_1 x_2 x_3 + %x_1 x_2 x_4
%                       \dotsb + x_{n-2} x_{n-1} x_n),
\end{align}
and so on
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this it can combinatorially be proved that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yRf$, $\yrs$.

@@

We need to make an initial probability assignment before any experimental
observations are made, no matter what kinds of predictions we are
interested in. This initial assignment will be modified by the experimental
observations.

We yet know very little about the physical details of the individual
neurons, their locations for example. Our initial state of knowledge $\yH$
is therefore symmetric, or \enquote{exchangeable}, under their
permutations. This symmetry must be reflected in our initial probability:
the \emph{representation theorem for finite exchangeability} states that it
must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \| \yH) = \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. Proof
and generalizations to non-binary and continuum cases are given by
de~Finetti \cite{definetti1959b}, Ericson \cite{ericson1976}, Diaconis
\cite{diaconis1977}, Heath \amp\ Sudderth \cite{heathetal1976}. This
theorem is intuitive: owing to symmetry, all states with $N\yRf$ active
neurons must have equal probabilities.

@@

The conditional probability $\ypp(\yrs,\yRf)$ connects the spaces of the
sample average $\yXf\in\set{0,\dotsc,N}$ and of the population average
$\yxs\in\set{0,\dotsc,n}$. It is a coarsening projector of any probability
$p$ for $\yXf$ onto a marginal probability $p_*$ for $\yxs$:
\begin{equation}
  \label{eq:projection_P}
  % \ypp\colon p(\yXf=\dotv) \to
  p_*(\yxs=\yrs)
  = \sum_{N\yRf=0}^N \ypp(\yrs\|\yRf) p(\yXf=\yRf).
\end{equation}
Therefore it also a pulls back expectations of functions $f$ of the sample
average $\yxs$ to expectations of functions $f^*$ the population average
$\yXf$:
\begin{equation}
  \label{eq:pullback_P}
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
  \begin{gathered}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
    \\
   \expeb{f(\yxs)}= \expeb{f^*(\yXf)}
    =
    \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
    =
    \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}

\begin{figure}[!t]
\centering
%\includegraphics[width=0.75\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.75\linewidth]{pop_sample_projection2.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1} \defs
  \ypp(\yrs\|\yRf)$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, for
example \fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of
\enquote{fuzzy identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$
and $\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix.
We thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities; they may miss important
features of the two probability distributions. If the distribution for the
population average $\yXf$ is bimodal, for example, the bimodality can be
lost in the distribution for the sample average $\yxs$, owing to the
coarsening effect of $\ypp(\yrs\|\yRf)$.

The approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. Upon observation
of a sample average, say $\yxxs=a$, the updated expectations for such
average in the population and in any new sample will usually be shifted
towards the observed value, as follows from Bayes's theorem and the
formulae above.

@@

We prefer not to answer the question that closed the preceding section,
because the optimal answer can only be given case by case, depending on the
inferences we are trying to make, on the assumptions needed and negligible,
and on our computational power. We would just like to stress that the
answer is not unique, and some answers may imply assumptions we were not
aware of.

The probability distribution for the sample average obtained by applying
\me\ at the population level has by construction a lower entropy than the
one obtained by application at the sample level. It says that some sample
states $\yx$ should be considered less likely than in the other
distribution. The lower entropy signals the presence of an additional
physical assumption behind the former distribution. Such assumption,
intuitively, is that the sample is part of a larger population, of which it
is representative. And this assumption may affect our predictions.
\mynote{It is a very natural assumption in the neuroscientific context and
  therefore we find it preferable. In fact, also physical neuronal network
  model add an external input to the neurons, mimicking their embedding in a
  larger network}

\iffalse
The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that
\fi
The possibility of using two different distributions is
not a physical contradiction. Similar situations arise in statistical
mechanics. It is known that if a system is described by a \me\ Gibbs state,
its subsystems need not be \cite{maesetal1999}. A quandary somehow similar
to ours also appears in the statistical description of an equilibrium state
at the end of a non-equilibrium process: we can describe our knowledge
about it either by a Gibbs distribution, or by the Liouville-evolved Gibbs
distribution associated with equilibrium state at the beginning of the
process. The two descriptions differ -- even though the final physical
state is exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The appearance
of this difference is natural: in one case we can make sharper predictions
about the state thanks to our knowledge of its preceding dynamics.
\mynote{but in this case both distributions are immensely sharp, and this
  doesn't affect predictions}

>>>>

%% Author: PGL  Porta Mana
%% Created: 2015-11-04T09:20:08+0100
%% Last-Updated: 2017-05-18T01:18:11+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Report-no: ***
\documentclass{article}

@@

\usepackage[final,nonatbib]{nips_2017}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\pdftitle}{Representative random samples\\ and maximum-entropy
  distributions:\\ a dilemma}
\newcommand*{\headtitle}{\pdftitle}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, V. Rostami, E. Torre}

@@

\author{
  P.G.L. Porta Mana\thanks{Equally contributing first authors}\\Independent researcher\\\texttt{pgl@portamana.org}
  \And
  V. Rostami$^*$\\Forschungszentrum J{\"u}lich INM-6\\Germany\\\texttt{v.rostami@fz-juelich.de}
  \And
E. Torre\\Moneylaunderingland  
  % David S.~Hippocampus\thanks{Use footnote for providing further
  %   information about author (webpage, alternative
  %   address)---\emph{not} for acknowledging funding agencies.} \\
  % Department of Computer Science\\
  % Cranberry-Lemon University\\
  % Pittsburgh, PA 15213 \\
  % \texttt{hippo@cs.cranberry-lemon.edu} \\
  %% examples of more authors
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \AND
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
  %% \And
  %% Coauthor \\
  %% Affiliation \\
  %% Address \\
  %% \texttt{email} \\
}

@@

These computational costs are one, probably not the earliest, of the many
reasons why \me\ methods have been introduced in neuroscience. It would be
useful to somehow compress the statistical wealth of large neuronal
recordings into few quantities, like sample moments for example. This
compression would also entail interesting physico-biological properties of
neuronal activity. The standard \me\ method
\cite{jaynes1957,sivia1996_r2006,meadetal1984} accomplishes this kind of
compression: it associates a unique probability distribution with few
experimental quantities \cite{***}. But this is only one of its uses. It is
also used for various information-theoretic purposes or to generate
reference probability distributions \cite{***}. In all these uses the \me\
distribution is chosen as the \enquote{maximally noncommittal} one
\cite{jaynes1963}. This adjective means little without further technical
characterizations. Different works give different characterizations, but in
this paper we will use the quoted expression as an umbrella term for all of
them. Our results will not depend on the specific characterization of
\enquote{noncommittal}.

@@

In the final discussion we present some arguments in favour of one choice
in our dilemma, but we do not mean that to be our final answer. This note,
and its dilemma,  also have the more general purpose of reminding our
readers the importance of asking \emph{what is the question we're trying to
  answer with probability theory?} and \emph{which assumptions do we want
  or need to make?}

Our mathematical analysis pertains to a neurons modelled as binary units at
one specific instant or short window in time. It is possible to similarly
discuss multi-state neuron models and population dynamics; but for
simplicity neurons are here assumed to be in a fixed, active \enquote{$1$}
or inactive \enquote{$0$} state; and evolution, change, time correlations,
and similar concepts do not concern us. We consider \me\ models that have
constraints based on some kinds of sample or population averages; they are
often called \enquote{homogeneous}. The final discussion touches upon
\enquote{inhomogeneous} models as well.

The notation in this note follows ISO and ANSI standards
\citep{iso1993,ieee1993,nist1995} but for the use of the comma \enquote{,}
to denote logical conjunction. Probability notation follows Jaynes
\citep{jaynes1994_r2003}. We intend \enquote{probability} as the degree of
belief which \enquote{would be agreed by all rational men if there were any
  rational men} \cite{good1966}.

@@

We have a population of $N$ binary neurons. We assume that they can be
distinguished, by their spike shapes for example; but other details, like
their locations, are unknown. The neurons have a joint state
$(X_1,\dotsc,X_N) \defs \yX$ having fixed but unknown binary values
$(R_1,\dotsc,R_N)\defs \yR \in \set{0,1}^N$. A particular sample of $n$
neurons from this population has joint state $(x_1, \dotsc, x_n) \defs \yx$
having fixed binary values $(r_1, \dotsc, r_n)\defs\yr$. We will consider
various averages of the population and the sample. For this purpose we
introduce a general averaging operator $\sav{\dotv}$ defined by
\begin{equation}
  \begin{gathered}
    \av{\yX} \defd \tfrac{1}{N} (X_1 + X_2 + \dotsb + X_N),
    % & \sav{\yx} &\defd \tfrac{1}{n} (x_1 + x_2 + \dotsb + x_n),
   \qquad% \\[2\jot]
    \av{\yX \yX} \defd \tbinom{N}{2}^{-1} (X_1 X_2 + X_1 X_3 + \dotsb +
    X_{N-1} X_N),
    % & \sav{\yx \yx} &\defd\tbinom{n}{2}^{-1} (x_1 x_2 + x_1 x_3 + \dotsb
    % + x_{n-1} x_n),
    \\[2\jot]
    \av{\yX\yX\yX} \defd \tbinom{N}{3}^{-1} (X_1 X_2 X_3 +% X_1 X_2 X_4
    \dotsb + X_{N-2} X_{N-1}X_N),
    % & \sav{\yx\yx\yx} &\defd \tbinom{n}{3}^{-1} (x_1 x_2 x_3
    % + %x_1 x_2 x_4
    % \dotsb + x_{n-2} x_{n-1} x_n),
  \end{gathered}
\end{equation}
and so on.
These formulae say that $\yXf$ is the fraction of active neurons, $\yXXf$
the fraction of simultaneously active pairs out of all $\binom{N}{2}$
pairs, $\av{\yX\yX\yX}$ the fraction of simultaneously active triplets, and
so on. Products of states like $X_i \dotsm X_j$ also have values in
$\set{0,1}$; from this we can combinatorially  prove that
\begin{equation}
  \label{eq:products_intermsof_average}
  \av{\underbrace{\yX\dotsm\yX}_{\text{$m$ factors}}}
  = \binom{N}{m}^{-1}\binom{N\yXf}{m}.
\end{equation}
Analogous formulae  hold for quantities like $\yx$, $\yR$, $\yr$.

@@

We need to make an initial probability assignment before any experimental
observations are made. %, no matter what kinds of predictions we are interested in.
This initial assignment will be modified by our experimental
observations.

We yet know very little about the physical details of the individual
neurons; their locations for example. Our initial state of knowledge $\yH$
is therefore symmetric, or \enquote{exchangeable}, under their
permutations. This symmetry must be reflected in our initial probability:
the \emph{representation theorem for finite exchangeability} states that it
must obey
\begin{equation}
  \label{eq:joint_plaus_N_homog}
  \p(\yX = \yR \| \yH) = \binom{N}{N\yRf}^{-1} \p(\yXf=\yRf \| \yH),
\end{equation}
the latter being the probability for the population average $\yX$. Proof of
this theorem and generalizations to non-binary and continuum cases are
given by de~Finetti \cite{definetti1959b}, Ericson \cite{ericson1976},
Diaconis \cite{diaconis1977}, Heath \amp\ Sudderth \cite{heathetal1976}.
This theorem is intuitive: owing to symmetry, all states with $N\yRf$
active neurons must have equal probabilities.

@@

The conditional probability $\ypp(\yrs,\yRf)$ relates the spaces of the
sample average $\yXf\in\set{0,\dotsc,N}$ and of the population average
$\yxs\in\set{0,\dotsc,n}$ in a special way. It is a coarsening projector of
any probability $p$ for $\yXf$ onto a marginal probability $p_*$ for
$\yxs$:
\begin{equation}
  \label{eq:projection_P}
  % \ypp\colon p(\yXf=\dotv) \to
  p_*(\yxs=\yrs)
  = \sum_{N\yRf=0}^N \ypp(\yrs\|\yRf) p(\yXf=\yRf).
\end{equation}
Conversely it also a pulls back expectations of functions $f$ of the sample
average $\yxs$ to expectations of functions $f^*$ of the population average
$\yXf$:
\begin{equation}
  \label{eq:pullback_P}
%  \ypp^*\colon f(\yxs=\dotv) \to f^*(\yXf=\dotv) \defd \sum_{N\yrs=0}^n
%  f(\yrs)\ypp(\yrs\|\dotv),
  \begin{gathered}
    f^*(\yXf) \defd \sum_{n\yrs=0}^n f(\yrs)\ypp(\yrs\|\yXf),
    \\
   \expeb{f(\yxs)}= \expeb{f^*(\yXf)}
    =
    \sum_{n\yrs=0}^n f(\yrs)\,\p(\yxs=\yrs \| \yH)
    =
    \sum_{N\yRf=0}^N f^*(\yRf)\,\p(\yXf=\yRf \| \yH).
  \end{gathered}
\end{equation}

\begin{figure}[!t]
\centering
%\includegraphics[width=0.75\linewidth]{pop_sample_projection.pdf}\\
\includegraphics[width=0.75\linewidth]{pop_sample_projection2.pdf}%
\caption{Log-plot of the hypergeometric distribution
  $%\p(\yxs = \yrs \|\yXf=\yRf, \yH)=
  \binom{n}{n\yrs}\binom{N-n}{N \yRf-n\yrs}\binom{N}{N \yRf}^{-1} \defs
  \ypp(\yrs\|\yRf)$ for $N=5000$, $n=200$. (Band artifacts may appear in the
  colourbar depending on your \textsc{pdf} viewer.)}
\label{fig:hypergeom_proj}
\end{figure}%hypergeometric_identity.nb
A look at a plot of the hypergeometric distribution $\ypp(\yrs\|\yRf)$, for
example \fig~\ref{fig:hypergeom_proj}, reveals that it is a sort of
\enquote{fuzzy identity matrix} between the $\yXf$-space $\set{0,\dotsc,N}$
and $\yxs$-space $\set{0,\dotsc,n}$. When $n=N$ it is the identity matrix.
We thus have that
\begin{equation}
  \label{eq:roughly_equal_nN}
  \p(\yxs=a) \approx \p(\yXf=a),\qquad
\expeb{f(\yxs)} \approx \expeb{f(\yXf)}.
\end{equation}
These are only very approximate equalities; they may miss important
features of the two probability distributions. In the next section we will
in fact emphasize their differences. If the distribution for the population
average $\yXf$ is bimodal, for example, the bimodality can be lost in the
distribution for the sample average $\yxs$, owing to the coarsening effect
of $\ypp(\yrs\|\yRf)$.

Yet the approximate equalities above express the fact that \emph{our
  uncertainty about the sample is representative of our uncertainty about
  the population and about other samples}, and vice versa. Upon observation
of a sample average, say $\yxxs=a$, the updated expectations for such
average in the population and in any new sample will usually be shifted
towards the observed value, as follows from Bayes's theorem and the
formulae above.

@@

The question that closed the preceding section cannot receive a categorical
answer. An answer can only be given case by case, depending on the
inferences we are trying to make, on the assumptions needed and those
negligible, and on our computational power. We want to stress, in fact, the
extreme importance and priority of the following two questions:
\begin{itemize}
\item Of which quantities do we need to quantify the uncertainty? and why?
\item What are the assumptions we need or want to make to answer the
  question above?
\end{itemize}

The first purpose of this note was to remind our readers that probability
models can have hidden assumptions that we don't want to make or that are
in contradiction with assumptions we want to make. The dilemma generated by
the \me\ method and the assumption of random representative sampling is an
example. The usual application of \me\ at the sample level generates a
different distribution than the application at the population level. Then
the former application must implicitly assume that no larger population
exist; or in other words that the neurons of the sample are isolated from
their environment.

The probability distribution for the sample average obtained by applying
\me\ at the population level has by construction a lower entropy than the
one obtained by application at the sample level. The former distribution
says that some sample states $\yx$ should be considered less likely than in
the latter distribution. A lower entropy signals the presence of an
additional assumption. Such assumption, intuitively, is that the sample is
part of a larger population, and our ignorance about the former is
representative of our ignorance about the latter. This assumption may
affect our predictions.

In a neuroscientific context we find this assumption very natural, and
therefore the \me\ method at the population level preferable. Also
\emph{physical} neuronal-network models add indeed an external input to the
neurons, mimicking their embedding in a larger network.

\iffalse
The quandary we found in the present note has at least two bright sides.
The first is that the two different \me\ applications may give numerically
similar results in some cases. One could thus be used as an approximation
of the other, depending on the circumstances -- \eg\ if probability tails
are unimportant. See \fig~\ref{fig:diff_maxent_pop_sample_realdata} for an
example.
\begin{figure}[!t]
\centering
\includegraphics[width=0.95\columnwidth]{different_maxent_pop_sample_200_realdata_both.pdf}
% 
\caption{Example of discrepancy in $\p(\yxs = \yrs \| \yH)$ constructed by
  \me\ at the population level followed by sample marginalization (blue
  triangles), and at the sample level (red circles). The constraints are
  the same in either case: $\expe{\yxs \|\yH}=0.045$,
  $\expe{\yxxs \|\yH}=0.0025$, and $N=5000$, $n=200$.}
\label{fig:diff_maxent_pop_sample_realdata}
\end{figure}%maxent_pop_or_sample.nb
The second is that
\fi
The possibility of using two different distributions is
not a physical contradiction. Similar situations arise in statistical
mechanics. It is known that if a system is described by a \me\ Gibbs state,
its subsystems need not be \cite{maesetal1999}. A quandary somehow similar
to ours also appears in the statistical description of an equilibrium state
at the end of a non-equilibrium process: we can describe our knowledge
about it either by a Gibbs distribution, or by the Liouville-evolved Gibbs
distribution associated with equilibrium state at the beginning of the
process. The two descriptions differ -- even though the final physical
state is exactly the same \cite[\sect~4]{jaynes1985d_r1993}. The appearance
of this difference is natural: in one case we can make sharper predictions
about the state thanks to our knowledge of its preceding dynamics.
%\mynote{but in this case both distributions are immensely sharp, and this
%  doesn't affect predictions}
